Measuring the double-drift illusion with hand trajectories

### 2021-12-11 Manuscript.Rmd
integrated Patrick's changes and feedback on Discussion

---
title: "Measuring the double-drift illusion with hand trajectories"
author: 
- Bernard Marius 't Hart 1,*
- Denise Y.P. Henriques 1
- Patrick Cavanagh 1,2,3
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document:
    reference_docx: template.docx
header-includes:
- \usepackage{lineno}
- \linenumbers
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{amsmath}
bibliography: bibliography.bib
csl: journal-of-vision.csl
link-citations: yes
---

@@

We examined 4 simple models. If there is a spatial offset limit that triggers resets, the reset points should fall on, or be close to a line with a constant x coordinate (blue line in Fig. 5A). If there is a temporal limit, all resets should have the same or a similar reset time (distance from origin / speed, the red arcs in Fig. 5A). If we assume thaat these spatial or temporal limits have some random variation from trial to trial, we can fit a normal distribution (with 2 parameters: mean and standard deviation) on both the x coordinates and the time points of resets. However, neither the x coordinates, nor the time points of resets were normally distributed (both _p_<.001, Kolmogorov-Smirnov tests of data against optimal normal distribution) suggesting that there is no hard space or time limit that triggers resets. We then used gamma distributions to capture any processes where the resets happen at random time in a trial or a random illusory offset. So we additionally fit gamma distributions (with 2 parameters: rate and shape) on the X coordinates and time points of resets as well (see Fig 5A, marginal distributions; Fig 5B,C). We then calculate the likelihoods of these four simple distribution models both using all resets and within individual participants and use this to calculate an AIC to determine a relative probability to compare the models.

```{r fig5, echo=FALSE, message=FALSE, fig.width=6, fig.height=5.333333, fig.cap="\\label{fig:fig5}**Figure 5:** Reset models. **A:** Model limits relative to reset point locations. Reset points neither fall along a spatial limit (blue line) or a temporal limit (red lines: continuous for 4 _s_ passes, dashed for 3 _s_ passes). Marginal distributions show that reset points time and spatial offset approximate a gamma distribution. **B,C:** Color illustrates the gamma probability distributions, where white is zero probability, and the darkest red or blue is the maximum probability density. This is a relatively scale-free unit, where the total surface of the full probability density function is 1. The data and likelihood distributions match fairly well. **B:** Gamma distribution on time (log likelihood: -766.6). **C:** Gamma distribution on spatial offset (log likelihood: -1319.3)."}
invisible(capture.output(plotModels()))
```

@@

We first tested whether manual tracking of a double-drift stimulus is susceptible to the illusion as reported by perceptual judgments, pointing [@Lisi2017], and memory saccades [@Massendari2018; @Ueda2018]. The results of this online tracking experiment showed a close agreement between the angle of the manual tracking and the illusion strength expected based on the model from earlier perceptual measures [@Heller2021]. However, we were also interested in spontaneous resets and only ~4% of trials in this first experiment had deviations that could have been a reset. In contrast, in the second experiment, 63.8% of trials showed a reset. There was a reset within 1 _s_ in 22.3% of trials, and within 2 _s_ in 52.2% of trials. We can use these results to determine the rate of spontaneous resets that would have been expected in the first experiment. If each pass is independent and the perceived trajectories were equivalent in the two experiments, this would mean ~98.5% of trials of Experiment 1 should have shown at least 1 spontaneous reset which is far larger than the ~4% of trials we observed ( _p_<.001, exact binomial test). We are not sure why so few spontaneous resets occurred in the first experiment. Possible explanations could be either that online tracking requires relatively fast hand movements, leaving little time for drawing the details of any spontaneous resets or the demands of the online tracking increased the attention to the double-drift stimulus and with less chance of attentional distraction, the likelihood of resets was reduced [@Nakayama2020]. The one feature that sets this task apart from others is that the tracking is done while perceiving the stimulus, such that action could affect perception. One possible mechanism is that predicted sensory consequences of the hand movement influence perception ("I will move my hand in this direction, so I expect to see motion in that direction as well"). And another is post-dictive inference ("my hand has moved in this direction so I must have seen the visual stimulus move in the same direction"). But this is a question for future research.

In our second experiment, we asked participants to draw the perceived path of the stimulus after observing a single pass. The orientation of the drawn traces reflect the strength of the illusion, and in a majority of trials (63.8%) there was a clearly indentifiable reset point, of either a "return" or a "hit-the-wall" type. We then set out to test if these spontaneous resets were triggered by a limiting distance of drift away from the physical location, or by a specific limit of time. Neither of these provided a viable explanation of the pattern of resets. Resets points are broadly distributed throughout the workspace, suggesting that resets are not linked to illusion strength or stimulus onset, but may caused by a more "random" process.

We observe that reset times appear randomly sampled from a broad gamma distribution. In part, this may rely on how (un)suitable hand trajectories are as an accurate measure of reset points, as they may contain an unknown amount of noise. Therefore, we compare our results with those from an as of yet unpublished work [pre-print: @Liu2021]. This study probed how spontaneous resets might affect the accumulated illusion throughout time (1, 2 or 3 _s_) while keeping the actual path (2 or 4 _dva_) and maximum illusion constant (45° deviation, so 1.4 or 2.8 _dva_ offsets, respectively). Spontaneous resets should decrease the accumulated illusion (see Fig 6A). After the main stimulus presentation, there is a 350 _ms_ gap, during which the double-drift stimulus is not shown, and this is know to evoke a reset. After this gap-evoked reset the illusion resumes accumulating from the veridical position where the gabor appears after the gap [@Lisi2015]. By changing the location where the double-drift reappears and asking participants if it jumped left or right, the point of subjective equivalence that expresses the size of the gap-evoked jump reset can be determined.

The data from 10 participants showed that for a fixed path and illusion strength, longer durations produced smaller illusory offsets at the end (Fig 6B, round dots). This can be explained by (among others) resets randomly occuring throughout time. Liu et al. [-@Liu2021] present a Poisson process model with a constant chance of reset occurring every 10 ms of ~0.007 or ~0.7 / _s_ (Fig 6B, light diamonds). If we compare this to the gamma distribution fit, with a rate of $2.81$ events per second, and reset occurring at the $3.74^{th}$ event (on average), this amounts to a ~0.75 (2.81/3.74) chance of a reset per second. If we simply take the parameters of the gamma distribution fit to reset times in the re-tracing experiment (without optimization), we can also predict PSEs (Fig 6B, dark diamonds). We see that the two findings, based on very different approaches, are mostly in agreement with each other.

```{r fig6, echo=FALSE, message=FALSE, fig.width=6, fig.height=4, fig.cap="\\label{fig:fig6}**Figure 6:** Spontaneous resets without hand trajectories. Experiment and Results based on data from Liu et al. [-@Liu2021], re-analyzed here. **A:** By ensuring the perceived path of a double-drift stimulus is straight down (teal arrows), any gap-induced reset, should be equal to the accumulated illusion if there are no spontaneous resets (black arrow), but smaller if there are (light blue arrow). Gaps were presented after 1, 2 or 3 seconds, and 2 or 4 dva real paths, which allows psychophysics to quantify the amount of illusion left after several periods of abserving a double-dirft stimulus. **B:** Points of subject equivalence show that the size of the illusion decreases with time even as the real path, and maximum illusion is kept constant. These data are in line with resets occurring randomly over time. The simplest model presented by Liu et al. (2021) keeps the illusion strength constant and fits a chance of reset per time unit as ~0.007 / 10 ms to match the gap-evoked reset (diamond symbols)."}
plotJumpPSEs(target = 'inline', individualDots = FALSE, modelDots = TRUE, gammaDots=TRUE)
```

Kwon, Tadin, and Knill [-@Kwon2015] proposed a Bayesian object tracking model to explain the double-drift illusion and other phenomena. In their model, when the position signals are weak, the motion signals bias perceived position away from the true location. Their model predicts that this deviation saturates after about 200 _ms_ and the perceived path then continues parallel to the physical path, as in the “hit the wall” resets we describe above [@Kwon2015, Fig. 3]. In contrast to their model, the reset in our re-tracing task show quite variable timing, and do not show a clear spatial offset (Fig. 5A) but seem to fall on a fairly wide gamma distribution on reset time (Fig. 5B), which is at odds with a set limit (either spatial or temporal) where saturation or resets occur. This seems to argue against a perceptual mechanism underlying resets, such as on going Bayesian inference until the discrepancy between retinal and perceived position signals is too large. In any case, the best fitting temporal limit of these reset points was not 200 _ms_ but 1.3 _s_ and 99% of detected reset times are above 200 _ms_ (see Fig. 5A). Moreover, 4/9 of our participants clearly reported “return” resets that the Kwon et al. [-@Kwon2015] model does not predict, and only 1/9 participants showed clear "hit-the-wall" resets. The “return” resets suggest that both the physical and perceived locations are available to the visual system and that when spontaneous resets occur, the perceived location returns toward the physical location. While our results speak against a particular, stimulus or perception induced mechanism leading to resets, the illusion itself may still have limits, and the way resets are perceived may relate to stimulus properties as well as to illusory perception. Regardless, future models that aim to predict perception in double-drift stimuli need to also predict resets.

Recently, it has been suggested that resets can be triggered by a distraction of attention [@Nakayama2020]. The participants in this experiment also retraced the perceived path and reported “return” resets. Nakayama and Holcombe [-@Nakayama2020] suggest that planned eye movements can trigger a “return” reset back to the physical path. In our experiment, we did not control for eye-movements, so we can't directly investigate whether eye movements accompanied the “return” resets that participants reported. However, the random distribution of resets throughout time are in line with a mechanism based on distracted attention, in some cases leading to eye-movement planning.

We find that manual tracking of the double-drift stimulus during its motion did show the expected illusion, although not always as strong. Few if any resets were detected during online tracking, perhaps because of the additional attentional requirements or the short duration (2 s), although interactions with the simultaneous hand movements could explain this too. Participants did report spontaneous resets in delayed, single-pass re-tracing of Experiment 2. The resets were best explained by a gamma distribution on reset time. This finding seems to indicate that spontaneous resets occur independently from the stimulus or how it is perceived, and spontaneous reset may even be random. If there is a mechanism causing resets we have not uncovered it here, and if resets are independent from perception, the illusory percept itself may still be limited, and give rise to how resets are perceived.

>>>>

---
title: "Measuring the double-drift illusion with hand trajectories"
author: 
- Bernard Marius 't Hart 1,*
- Denise Y.P. Henriques 1
- Patrick Cavanagh 1,2,3
output:
  html_document:
    df_print: paged
  word_document:
    reference_docx: template.docx
  pdf_document: default
header-includes:
- \usepackage{lineno}
- \linenumbers
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{amsmath}
bibliography: bibliography.bib
csl: journal-of-vision.csl
link-citations: yes
---

@@

We examined 4 simple models. If there is a spatial offset limit that triggers resets, the reset points should fall on, or be close to a line with a constant x coordinate (blue line in Fig. 5A). If there is a temporal limit, all resets should have the same or a similar reset time (distance from origin / speed, the red arcs in Fig. 5A). If we assume that these spatial or temporal limits do allow a little random variation from trial to trial, we can fit a normal distribution (with 2 parameters: mean and standard deviation) on both the x coordinates and the time points of resets. However, neither the x coordinates, nor the time points of resets were normally distributed (both _p_<.001, Kolmogorov-Smirnov tests of data against optimal normal distribution). Additionally, if resets are triggered by either a spatial or temporal limit the data would all fall relatively close to it, but this does not appear to be the case.

These first two models assumed that resets were triggered at fixed locations in space or time, but the next tro propose that the resets are random processes that occur with some distribution across space or time. We used gamma distributions to capture these processes (with 2 parameters: rate and shape) on the X coordinates and time points of resets as well (see Fig 5A, marginal distributions; Fig 5B,C). To compare across the four models, we then calculated the likelihoods of their predictions (based on the underlying distributions' probability density functions) using all resets together and also within individual participants. From these likelihood values, we calculated an AIC to determine a relative probability for each model.

```{r fig5, echo=FALSE, message=FALSE, fig.width=6, fig.height=5.333333, fig.cap="\\label{fig:fig5}**Figure 5:** Reset models. **A:** Model limits relative to reset point locations. Reset points neither fall along a spatial limit (blue line; log likelihood: -1415.4) or a temporal limit (red lines: continuous for 4 _s_ passes, dashed for 3 _s_ passes; log likelihood: -837.0). Marginal distributions show that reset points time and spatial offset approximate a gamma distribution. **B,C:** Color illustrates the gamma probability distributions, where white is zero probability, and the darkest red or blue is the maximum probability density. This is a relatively scale-free unit, where the total surface of the full probability density function is 1. The data and likelihood distributions match fairly well. **B:** Gamma distribution on time (log likelihood: -766.6). **C:** Gamma distribution on spatial offset (log likelihood: -1319.3)."}
invisible(capture.output(plotModels()))
```

@@

We first tested whether manual tracking of a double-drift stimulus is susceptible to the illusion as reported by perceptual judgments, pointing [@Lisi2017], and memory saccades [@Massendari2018; @Ueda2018]. The results of this online tracking experiment showed a close agreement between the angle of the manual tracking and the illusion strength expected based on the model from earlier perceptual measures [@Heller2021]. However, we were also interested in spontaneous resets and only ~4% of trials in this first experiment had deviations that could have been a reset. In contrast, in the second experiment, 63.8% of trials showed a reset. There was a reset within 1 _s_ in 22.3% of trials, and within 2 _s_ in 52.2% of trials. We can use these results to determine the rate of spontaneous resets that would have been expected in the first experiment if it had the same reset dynamics as the second experiment. If each pass was independent and the perceived trajectories were equivalent in the two experiments, this would mean ~98.5% of trials of Experiment 1 should have shown at least 1 spontaneous reset which is far larger than the ~4% of trials we observed ( _p_<.001, exact binomial test). We are not sure why so few spontaneous resets occurred in the first experiment. We first suggest two simple explanations for the difference: 1) Online tracking required relatively fast hand movements, and therefore left little time for drawing the details of any spontaneous resets. 2) The demands of the online tracking increased the attention to the double-drift stimulus and with less chance of attentional distraction, the likelihood of resets was reduced [@Nakayama2020]. However, this task, unlike previous tests of the double-drift stimulus, may involve an interaction between action and perception. For example, the predicted sensory consequences of the hand movement may influence perception ("I will move my hand in this direction, so I expect to see motion in that direction as well"). Or vice versa, as in post-dictive inference ("my hand has moved in this direction so I must have seen the visual stimulus move in the same direction"). But this is a question for future research.

In our second experiment, we asked participants to draw the perceived path of the stimulus after observing a single pass. The orientation of the drawn traces reflect the strength of the illusion, and in a majority of trials (63.8%) there was a clearly indentifiable reset point, of either a "return" or a "hit-the-wall" type. We then set out to test if these spontaneous resets were triggered by a limiting distance of drift away from the physical location, or by a specific limit of time. Neither of these provided a viable explanation of the pattern of resets. Resets points are broadly distributed, suggesting that resets are not linked to illusion strength or stimulus onset, but may be caused by a more "random" process.

Recently, it has been suggested that resets can be triggered by a distraction of attention [@Nakayama2020]. The participants in their experiment also retraced the perceived path and reported “return” resets. Nakayama and Holcombe [-@Nakayama2020] claim that planned eye movements can trigger a “return” reset back to the physical path. In our experiment, we did not control for eye-movements, so we can't directly investigate whether eye movements accompanied the “return” resets that participants reported. However, the random distribution of resets throughout time are in line with a mechanism based on distracted attention, which may involve eye-movement planning and execution. The rate of resets that we found here, ~0.75 / _s_, are too infrequent for microsaccades (~3 / _s_) so the distractions may be more significant than these small eye movements. On the other hand, random occurrance of distractions, a fraction of which triggers a reset, is in line with the kind of processes underlying gamma distributions [e.g. @McGill1965; @Wolfe2002].

We observe that reset times were well modeled by broad gamma distribution, suggesting that resets were best explained as occurring randomly in time with an approximate rate of 1 every 1.3 _s_. However, we have to qualify our conclusions given the noisy nature of hand-drawn trajectories as a measure of the perceived trajectory and the reset points. Therefore, we compare our results with those from s similar study that analyzed trajectory endpoints [pre-print: @Liu2021]. In this study, the internal speed of the Gabor texture was adjusted across conditions and participants so that the illusion always had a local direction of 45° compared to its actual vertical direction. This then generated a highly predictable offset of the end of the trajectory that would be $\sqrt{2}$ of the path length if there were no spontaneous resets. Anything less than that revealed the presence and size of spontaneous resets. Liu et al. tested 3 different durations (1, 2 and 3 s) and two path lengths (2 and 4 dva) for the first segment of the Gabor’s motion (see Fig. 6A). They then presented a 350 _ms_ gap during which the double-drift stimulus is not shown, and this is known to evoke a full reset to the physical location [@Lisi2015]. After this gap-evoked reset the illusion resumes accumulating from the veridical position where the gabor appears after the gap [@Lisi2015]. By changing the location where the double-drift reappeared and asking participants if it shifted left or right, the endpoint of the first segment was determined. The difference between that point and the expected value for no spontaneous resets indicated the presence and size of any that had occurred.

The data from 10 participants showed that for a fixed path and illusion strength, longer durations produced smaller illusory offsets at the end (Fig 6B, round dots). This can be explained by (among others) resets randomly occuring throughout time. Liu et al. [-@Liu2021] present a Poisson process model with a constant chance of reset of ~0.7 / _s_ (Fig 6B, light diamonds). In comparison, with the gamma distribution fit to our data here (rate=$2.81$, shape=$3.74$), we found a ~0.75 / _s_ chance of a reset. Moreover, if we simply take the parameters of the gamma distribution fit to reset times in the re-tracing experiment (without further optimization), we can also generate the PSEs that should be found in Liu et al.'s experiment (Fig 6B, dark diamonds). We see that the two findings, based on very different approaches, are mostly in agreement with each other.

```{r fig6, echo=FALSE, message=FALSE, fig.width=6, fig.height=4, fig.cap="\\label{fig:fig6}**Figure 6:** Spontaneous resets without hand trajectories. Experiment and Results based on data from Liu et al. [-@Liu2021], re-analyzed here. **A:** By ensuring the perceived path of a double-drift stimulus is straight down (teal arrows), any gap-induced reset, should be equal to the accumulated illusion if there are no spontaneous resets (black arrow), but smaller if there are (light blue arrow). Gaps were presented after 1, 2 or 3 seconds, and 2 or 4 dva real paths, which allows psychophysics to quantify the amount of illusion left after several periods of observing a double-drift stimulus. **B:** Points of subject equivalence show that the size of the illusion decreases with time even as the real path, and maximum illusion is kept constant. These data are in line with resets occurring randomly over time. The simplest model presented by Liu et al. (2021) keeps the illusion strength constant and fits a chance of reset per time unit as ~0.007 / 10 ms to match the gap-evoked reset (diamond symbols)."}
plotJumpPSEs(target = 'inline', individualDots = FALSE, modelDots = TRUE, gammaDots=TRUE)
```

Kwon, Tadin, and Knill [-@Kwon2015] proposed a Bayesian object tracking model to explain the double-drift illusion and other phenomena. In their model, when the position signals are weak they have less weight and the motion signals make a prediction that moves the perceived position away from the true location. Their model predicts that after about 200 _ms_ the motion prediction saturates and the perceived path then continues parallel to the physical path, as in the “hit the wall” resets we describe above [@Kwon2015, Fig. 3]. In contrast to their model, the resets in our re-tracing task show quite variable timing (Fig. 5A) and are better described by a fairly broad gamma distribution (Fig. 5B). This argues against a  mechanism that is based on comparing the current location to the physical location and saturating or triggering a reset when the discrepancy is too large. In any case, the best fitting temporal limit of the reset points in our data was not 200 _ms_ but 1.3 _s_ and 99% of detected reset times are above 200 _ms_ (see Fig. 5A). Moreover, 4/9 of our participants clearly reported “return” resets that the Kwon et al. [-@Kwon2015] model does not predict, and only 1/9 participants showed clear "hit-the-wall" resets. The “return” resets suggest that both the physical and perceived locations are available to the visual system and that when spontaneous resets occur, the perceived location returns toward the physical location. Our results suggest that the resets, whether saturation or returns, happen randomly over time, rather than at a particular deviation in spatial offset or after a fixed duration. Future models of position perception for the double-drift stimulus need to include a realistic model of the locations and type of resets.

In summary, we find that manual tracking of the double-drift stimulus during its motion did show the expected illusion at the expected strength (experiment 1). Few if any resets were detected during online tracking, perhaps because of the additional attentional requirements or the short duration (2 _s_), although interactions with the simultaneous hand movements could explain this too. In contrast, participants did report relatively frequent, spontaneous resets in the delayed, single-pass re-tracing of Experiment 2. The resets were best explained by a gamma distribution on reset time. This finding suggests that spontaneous resets occur independently of the stimulus and how or where it is perceived. Instead, they may occur randomly over time, perhaps due to unrelated attentional distractions.
